---
title: "catcheR overview: complete native pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{catcheR overview: complete native pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`catcheR` is an R package for CATCHER / iPS2-seq workflows. It contains functions for plasmid-library quality control, single-cell barcode catching, caught-cell filtering, no-catch/empty-cell handling, and barcode-swap correction.

This vignette is a map of the full workflow. The detailed usage for each module is described in the dedicated vignettes:

- `vignette("catcheR-step1QC")`
- `vignette("catcheR-step2QC")`
- `vignette("catcheR-10Xcatch")`
- `vignette("catcheR-filter-nocatch")`
- `vignette("catcheR-sortcatch")`

# Recommended project layout

A useful layout is to keep raw/intermediate files in `scratch/` and to write all package outputs to dated folders inside `Output/`.

```text
project/
  scratch/
    CATCHER1/
    CATCHER2/
    10X/
  Output/
    step1QC/
    step2QC_extraction/
    step2QC_plasmidQC/
    step2QC_combinedQC/
    10Xcatch/
    10XcatchQC/
    filtercatch/
    nocatch/
    sortcatch/
```

Each catcheR function writes a dated run folder:

```text
Output/<function>/YYMMDD_<function>_<sample>/
  csv/
  plots/
  stats/
  R_objects/
  to_scratch/
```

The `to_scratch/` folder contains only the files needed by the next step. You can copy them into `scratch/` if you want to restart later without rerunning previous steps.

# Full workflow skeleton

```{r full-workflow}
library(catcheR)

# Step 1: initial plasmid/intermediate QC
step1 <- catcheR_step1QC(
  folder      = "scratch/step1/",
  fastq.read1 = "plasmid_intermediate_R1.fastq.gz",
  output_dir  = "Output/",
  sample_name = "CATCHER1",
  DIs         = 100,
  ratio       = 10
)

# Step 2A: extract barcode/UCI/read-count information from final plasmid FASTQ
extract <- catcheR_step2QC_extraction(
  fastq      = "scratch/CATCHER1/final_plasmid_R1.fastq.gz",
  barcodes   = "scratch/CATCHER1/rc_barcodes_genes.csv",
  output_dir = "scratch/CATCHER1/results/",
  DIs        = 300
)

# Step 2B: single-library plasmid QC
qc1 <- catcheR_step2QC_plasmidQC(
  results_dir      = "scratch/CATCHER1/results/",
  output_dir       = "Output/",
  sample_name      = "CATCHER1",
  DIs              = 300,
  transfect_clones = 100,
  transfect_cells  = 2e6,
  nucleofect_cells = 2e6
)

# Optional: compare two independently generated libraries
combined <- catcheR_step2QC_combinedQC(
  lib1_dir         = "scratch/CATCHER1/results/",
  lib2_dir         = "scratch/CATCHER2/results/",
  output_dir       = "Output/",
  sample_name      = "CATCHER1_CATCHER2",
  lib1_name        = "CATCHER1",
  lib2_name        = "CATCHER2",
  DIs              = 300,
  transfect_clones = 100,
  transfect_cells  = 2e6,
  nucleofect_cells = 2e6
)

# Step 3: catch shRNA/barcode information in 10X reads
catch <- catcheR_10Xcatch(
  folder            = "scratch/10X/",
  fastq.read1       = "sample_R1.fastq.gz",
  fastq.read2       = "sample_R2.fastq.gz",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  samples           = 1,
  mode              = "bimodal"
)

# Step 4: inspect 10X catching QC and derive UMI threshold
catch_qc <- catcheR_10XcatchQC(
  folder      = "scratch/10X/",
  output_dir  = "Output/",
  sample_name = "AB017",
  samples     = 1,
  mode        = "bimodal"
)

# Step 5: filter caught cells
filtered <- catcheR_filtercatch(
  folder            = "scratch/10X/",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  output_dir        = "Output/",
  sample_name       = "AB017",
  samples           = 1
)

# Step 6: identify no-catch/empty cells if required
nocatch <- catcheR_nocatch(
  folder            = "scratch/10X/",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  output_dir        = "Output/",
  sample_name       = "AB017",
  samples           = 1
)

# Step 7: correct barcode swaps
sorted <- catcheR_sortcatch(
  folder      = "scratch/10X/",
  matrix      = "silencing_matrix_complete_1.rds",
  swaps       = "reliable_clones_swaps_DIs100_ratio10.csv",
  output_dir  = "Output/",
  sample_name = "AB017"
)
```

# Returned objects

Every high-level function returns a named list and also saves the same information to `R_objects/`.

```{r returned-objects}
names(qc1)
qc1$paths
qc1$plots
qc1$summary

# Reload a completed run later
qc1_reloaded <- readRDS("Output/step2QC_plasmidQC/YYMMDD_step2QC_plasmidQC_CATCHER1/R_objects/step2QC_plasmidQC_result.rds")
```

# Checkpoint and restart

To restart from a completed step, copy the contents of that run's `to_scratch/` folder into your scratch area:

```bash
cp Output/step2QC_plasmidQC/YYMMDD_step2QC_plasmidQC_CATCHER1/to_scratch/* scratch/CATCHER1/
```

Then use the scratch folder as input to the next function.
