---
title: "catcheR 10Xcatch workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{catcheR 10Xcatch workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Purpose

`catcheR_10Xcatch()` searches single-cell sequencing reads for CATCHER barcode information and links those barcode assignments to the expression matrix generated by Cell Ranger or a compatible single-cell preprocessing workflow.

This package version runs the bundled scripts directly. It does not require Docker.

# Required input files

The input folder should contain:

```text
scratch/10X/
  sample_R1.fastq.gz
  sample_R2.fastq.gz
  rc_barcodes_genes.csv
  filtered_feature_bc_matrix.h5   # or the expression matrix format expected by your script
```

The exact expression matrix file depends on how the upstream single-cell pipeline was run. The function keeps the original `expression.matrix` argument so that you can pass Cell Ranger output or another compatible matrix file.

# Run the complete 10X catching step

```{r run-10xcatch}
library(catcheR)

catch <- catcheR_10Xcatch(
  folder            = "scratch/10X/",
  fastq.read1       = "sample_R1.fastq.gz",
  fastq.read2       = "sample_R2.fastq.gz",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  reference         = "GGCGCGTTCATCTGGGGGAGCCG",
  UCI.length        = 6,
  threads           = 2,
  percentage        = 15,
  mode              = "bimodal",
  ratio             = 5,
  samples           = 1,
  x                 = 100,
  y                 = 400
)
```

# Main arguments

- `folder`: input directory containing FASTQ files, the expression matrix, and `rc_barcodes_genes.csv`.
- `fastq.read1`, `fastq.read2`: paired FASTQ files.
- `expression.matrix`: expression matrix file name inside `folder`.
- `reference`: reference sequence used by the catching script.
- `UCI.length`: expected UCI length.
- `threads`: number of threads passed to the shell script.
- `percentage`: percentage threshold used by filtering logic.
- `mode`: thresholding mode, usually `"bimodal"` or `"noise"`.
- `ratio`: ratio threshold.
- `samples`: number of samples to process.
- `x`, `y`: plotting/threshold parameters used by the underlying scripts.

# Outputs

`catcheR_10Xcatch()` is the complete catching pipeline. Internally, it calls the bundled shell script `barcode_silencing_slicing.sh`, which in turn calls the bundled R scripts for exploratory QC, cell filtering, empty-cell selection, and all-sample merging.

Typical outputs are written inside the input folder by the underlying scripts and can include per-sample `Results_*` folders, UMI-threshold files, intermediate caught-cell tables, and silencing matrices.

# Returned object

```{r inspect-10xcatch}
names(catch)
catch$paths
catch$runtime_seconds
```

# Modular alternative

If you prefer to run the pipeline step by step, use:

```{r modular-10x}
catch_qc <- catcheR_10XcatchQC(
  folder      = "scratch/10X/",
  output_dir  = "Output/",
  sample_name = "AB017",
  samples     = 1,
  mode        = "bimodal"
)

filtered <- catcheR_filtercatch(
  folder            = "scratch/10X/",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  output_dir        = "Output/",
  sample_name       = "AB017",
  samples           = 1
)

nocatch <- catcheR_nocatch(
  folder            = "scratch/10X/",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  output_dir        = "Output/",
  sample_name       = "AB017",
  samples           = 1
)
```

See `vignette("catcheR-filter-nocatch")` for details.
