---
title: "CATCHER workflow: from FASTQ to QC report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CATCHER workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

## Overview

The `catcheR` package provides an end-to-end R interface for analysing
shRNA plasmid library sequencing data. The four main functions mirror the
four steps of the protocol:

| Step | Function | Language |
|------|----------|----------|
| 1. Extract UMI/BC/UCI from FASTQ | `catcheR_step2QC_extraction()` | Calls Python |
| 2. Single-library QC | `catcheR_step2QC_plasmidQC()` | R |
| 3. Two-library comparison | `catcheR_step2QC_combinedQC()` | R |
| 4. Publication plots | `catcheR_step2QC_plots()` | R |

## Installation

```{r install}
# Install dependencies
install.packages(c("ggplot2", "dplyr", "tidyr", "scales",
                   "forcats", "patchwork", "data.table"))

# Install catcheR from a local folder
devtools::install("path/to/catcheR")

# Or from GitHub (once published)
# devtools::install_github("your-lab/catcheR")
```

## Step 1 — Extract clones from FASTQ

```{r step1}
library(catcheR)

catcheR_step2QC_extraction(
  fastq      = "EB003CS1_KS7_1_finalQC_S1_L001_R1_001.fastq",
  barcodes   = "rc_barcodes_genes.csv",
  output_dir = "CATCHER1_new/results/",
  DIs        = 300
)
```

`catcheR_step2QC_extraction()` calls the bundled Python script
`plasmid_final_corrected.py`, which extracts the barcode (pos 34–41)
and UCI (pos 42–47) from every read, matches barcodes against the
reference CSV, and writes `distribution_all_clones.csv` plus supporting
files to `output_dir`.

## Step 2 — Single-library QC

```{r step2}
qc <- catcheR_step2QC_plasmidQC(
  results_dir      = "CATCHER1_new/results/",
  DIs              = 300,
  transfect_clones = 100,       # clones from your test nucleofection
  transfect_cells  = 2000000,   # cells used in that test
  nucleofect_cells = 2000000    # cells planned for actual experiment
)
```

### Key outputs

- `DIs_threshold_stats_DIs300.txt` — read-count breakdown
- `clone_frequency_dist_DIs300.pdf` — log histogram separating noise from
  real clones
- `shrna_representation_3panel_DIs300.pdf` — UCI count distribution across
  shRNAs
- `calc_cells_and_dup_DIs300.pdf` — **central decision plot**: for each
  target m (distinct UCIs per shRNA), how many cells must you nucleofect
  and what is the expected duplication rate?
- `UCI_birthday_analysis_DIs300.txt` — full birthday-problem report

### Inspect results programmatically

```{r inspect}
# Per-shRNA data frame
head(qc$shrna)

# Birthday results table
qc$birthday$results

# Expected K at planned nucleofection
qc$birthday$K_planned
```

## Step 3 — Combined two-library QC

```{r step3}
combined <- catcheR_step2QC_combinedQC(
combined <-  catcheR_step2QC_combinedQC(
  lib1_dir         = "CATCHER1_new/results/",
  lib2_dir         = "CATCHER2_new/results/",
  out_dir          = "combined_results/",
  lib1_name        = "CATCHER1",
  lib2_name        = "CATCHER2",
  DIs              = 300,
  transfect_clones = 100,
  transfect_cells  = 2000000,
  nucleofect_cells = 2000000
)

# Check the recommendation
combined$overall_go   # TRUE = GO, FALSE = NO-GO
```

The go/no-go decision is based on two criteria:

1. **Reproducibility**: Spearman rho of UCI counts >= 0.80
2. **shRNA overlap**: >= 80% of shRNAs present in both libraries

## Step 4 — Publication plots

```{r step4}
catcheR_step2QC_plots(
  results_dir = "CATCHER1_new/results/",
  DIs         = 300
)
```

Produces pie charts, bar charts, and density plots of clone and shRNA
representation from the Python pipeline outputs.

## Understanding the birthday problem

The key calculation in `catcheR_step2QC_plasmidQC()` and `catcheR_step2QC_combinedQC()` is:

> If a shRNA has **n** distinct UCIs in the library and you recover **K**
> clones for that shRNA, some clones will share a UCI (duplicates).
> 
> E[distinct] = n × (1 − (1 − 1/n)^K)
> 
> % duplicates = (K − E[distinct]) / K × 100

**K is not chosen by you.** It follows from how many cells you nucleofect:

```
K = (cells_nucleofected × efficiency) / n_shRNAs
```

The `calc_cells_and_dup` plot shows, for each target **m** (distinct
UCIs you want per shRNA), how many cells you need and what % will be
duplicates. For CATCHER2 (mean n = 8.91):

| m | Cells needed | % dup | Verdict |
|---|---|---|---|
| 2 | 4M | 5.6% | OK |
| 3 | 5.9M | 10.8% | OK |
| 4 | 7.9M | 15.6% | OK |
| 5 | 9.9M | 20.1% | Borderline |
