---
title: "catcheR sortcatch workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{catcheR sortcatch workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Purpose

`catcheR_sortcatch()` performs barcode-swap correction on the final silencing matrix. It uses the swap table generated during plasmid/intermediate QC and the barcode reference file to correct assigned clone identities.

# Required input files

The input folder should contain:

```text
scratch/10X/
  silencing_matrix_complete_1.rds
  reliable_clones_swaps_DIs100_ratio10.csv
  rc_barcodes_genes.csv
```

The swap file is usually produced by `catcheR_step1QC()` and copied from its `to_scratch/` folder.

# Run sortcatch

```{r run-sortcatch}
library(catcheR)

sorted <- catcheR_sortcatch(
  folder      = "scratch/10X/",
  matrix      = "silencing_matrix_complete_1.rds",
  swaps       = "reliable_clones_swaps_DIs100_ratio10.csv",
  output_dir  = "Output/",
  sample_name = "AB017"
)
```

# Outputs

```text
Output/sortcatch/YYMMDD_sortcatch_AB017/
  csv/
    silencing_matrix_updated.csv
  R_objects/
    sortcatch_result.rds
    corrected_matrix.rds
  to_scratch/
    silencing_matrix_updated.csv
```

# Use the corrected matrix in R

```{r inspect-sortcatch}
names(sorted)
corrected_matrix <- sorted$corrected_matrix

dim(corrected_matrix)
head(colnames(corrected_matrix))
```

# Reload a previous run

```{r reload-sortcatch}
sorted <- readRDS("Output/sortcatch/YYMMDD_sortcatch_AB017/R_objects/sortcatch_result.rds")
corrected_matrix <- sorted$corrected_matrix
```

# Continue downstream

The corrected matrix in `to_scratch/` can be used as input for downstream single-cell analyses or copied into your project `scratch/` folder:

```bash
cp Output/sortcatch/YYMMDD_sortcatch_AB017/to_scratch/silencing_matrix_updated.csv scratch/downstream/
```
