---
title: "Colony PCR Sanger sequencing QC with catcheR_sangerQC()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colony PCR Sanger sequencing QC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

## Where this step fits in the CATCHER workflow

The CATCHER cloning strategy proceeds in two stages:

1. **Intermediate plasmid** — the shRNA hairpin, poly-T tail, and barcode
   (BC) are assembled into the intermediate plasmid backbone. At this stage
   the read contains *both* the shRNA sequence *and* the BC, which makes it
   possible to confirm that each BC is linked to the intended shRNA.

2. **Final plasmid** — the intermediate insert is transferred into the final
   AAV/lentiviral backbone. After this step only the BC is accessible by
   Sanger sequencing from the standard colony-PCR primers; the shRNA
   sequence is confirmed indirectly through the BC.

`catcheR_sangerQC()` is used **at the intermediate plasmid stage**, before
the final cloning step is complete. You pick bacterial colonies from the
ligation plate, run colony PCR, and send the PCR product for Sanger
sequencing. This function reads the resulting FASTA file and checks, for
each well:

- Was the insert present and readable?
- Does the barcode match the expected BC for that well?
- Does the shRNA sense/antisense sequence match the design?
- Is the shRNA hairpin structurally valid (sense = reverse complement of
  antisense)?
- Were there any point mutations introduced during synthesis or cloning?
- What was the sequencing quality in the BC and shRNA regions?

This QC step tells you whether your ligation worked correctly **before** you
invest time completing the cloning. It is especially important when:

- You want to confirm that the BC and shRNA are correctly paired (no
  barcode–shRNA swaps introduced during ligation).
- You are comparing two ligation ratios (e.g. 1:12 vs 1:20 insert:vector)
  and want to know which gives a higher correct-insert rate.
- You want to detect any point mutations introduced during oligo synthesis
  or PCR before propagating the plasmid library.

---

## What the sequencing read contains

The intermediate plasmid is sequenced with the primer **p-sOPTiKD_SEQ-FW**
(also called **siKDbc-seq_REV**), which reads the bottom strand 5'→3'. The
resulting read therefore contains the molecular identifiers in this order,
left to right:

```
5'(read) ── [vector] ── ACGCGT(MluI) ── [BC 8bp] ── [UCI 6bp] ──
           GGCGCGCC·ATTTAAAT·GTCGAC(anchor) ── AAAAAAA(polyA) ──
           [sense 21-22bp] ── CTCGAG(XhoI) ── [antisense 21-22bp] ──
           [vector] ── 3'(read)
```

Key points:

- The **BC** in the read is the reverse complement of the BC sequence in
  the design TOP OLIGO. `catcheR_sangerQC()` handles this automatically.
- The **sense** and **antisense** regions appear in the same order as in
  the design shRNA field — sense before the loop (CTCGAG), antisense after.
- A valid shRNA hairpin requires `sense == RC(antisense)`. The function
  checks this for every read and flags reads that fail as `INVALID HAIRPIN`.

---

## Anchor-based extraction

`catcheR_sangerQC()` does not rely on fixed read positions. Instead it
searches for the MCS anchor sequence:

```
GGCGCGCCATTTAAATGTCGAC
```

This is the reverse complement of `GTCGACATTTAAATGGCGCGCC` (SalI +
SwaI + AscI) on the top strand. Up to 2 mismatches are allowed in the
anchor search, which makes the extraction robust to occasional sequencing
errors in the vector backbone.

Once the anchor is located:

- **BC**: 8 bp immediately to the left of the anchor in the read.
- **UCI**: 6 bp to the left of the BC.
- **Sense**: sequence between the end of the anchor (which closes with
  GTCGAC) and the CTCGAG (XhoI/loop), after stripping leading A's from the
  poly-A tail.
- **Antisense**: 21–22 bp immediately after CTCGAG.

---

## Required inputs

| File | Description |
|------|-------------|
| FASTA file | Sanger sequences, one entry per well. Headers must end with the well ID preceded by `_` (e.g. `>KYO483_73984833_A1`). |
| Design Excel | Plate layout Excel file. Must contain a sheet (default `"Plate 01"`) with columns for well position, oligo name, and the full TOP OLIGO sequence. |
| ABI/PHD ZIP (optional) | ZIP archive of ABI output files including `.phd.1` quality files. When provided, the function reports minimum and mean Phred quality scores in the BC and shRNA regions. |

The design Excel is the same file used to order the oligos. `catcheR_sangerQC()`
parses the TOP OLIGO sequence to extract the expected BC and shRNA for each
well, so no separate barcode reference file is needed.

---

## Basic usage

```{r basic}
library(catcheR)

result <- catcheR_sangerQC(
  fasta        = "11109801799-1.fasta",
  design_xlsx  = "EB003C_DES01_shRNA_tdark_Ligation.xlsx",
  output_dir   = "Output/",
  sample_name  = "plate1_ligation",
  plate_sheet  = "Plate 01",
  abi_zip      = "11109801799-1_SCF_SEQ_ABI.zip"  # optional but recommended
)
```

---

## Comparing ligation conditions

If your plate is divided between two ligation ratios (e.g. columns 1–6 =
1:12, columns 7–12 = 1:20), set `ligation_split` accordingly. The function
reports pass rates separately for each condition.

```{r ligation}
result <- catcheR_sangerQC(
  fasta          = "11109801799-1.fasta",
  design_xlsx    = "EB003C_DES01_shRNA_tdark_Ligation.xlsx",
  output_dir     = "Output/",
  sample_name    = "plate1_1-12_vs_1-20",
  plate_sheet    = "Plate 01",
  ligation_split = 6L,                    # cols 1-6 = condition A
  ligation_labels = c("1:12", "1:20"),    # names for the two conditions
  abi_zip        = "11109801799-1_SCF_SEQ_ABI.zip"
)
```

The summary printed at the end shows pass rates per condition:

```
  1:12: 17/48 pass  |  1:20: 11/48 pass
```

---

## Inspecting results in R

The function returns a named list invisibly. All results are also
auto-saved as `.rds` files in `R_objects/`.

```{r inspect}
# Full per-well result table
result$results

# Status counts
result$summary

# Show only passing wells
result$results[grepl("^PASS", result$results$status), ]

# Show wells with mutations in the BC
result$results[!is.na(result$results$bc_mutations) &
               result$results$bc_mutations != "", ]

# Show wells where hairpin is invalid
result$results[result$results$hairpin_ok == "FALSE", ]

# Reload a previous run without re-running
result <- readRDS(
  "Output/sangerQC/260518_sangerQC_plate1_ligation/R_objects/sangerQC_result.rds"
)
```

---

## Output files

All output is written to:

```
Output/sangerQC/YYMMDD_sangerQC_<sample_name>/
  csv/
    sanger_QC_results.csv       # full per-well result table
  stats/
    sanger_QC_summary.txt       # status counts and extraction parameters
  R_objects/
    sangerQC_result.rds         # full result list (reloadable)
  to_scratch/
    sanger_QC_results.csv       # copy for downstream use
```

The CSV contains one row per Sanger well with these columns:

| Column | Description |
|--------|-------------|
| `sanger_well` | Well ID on the Sanger plate (A1–H12) |
| `ligation` | Ligation condition for this well |
| `seq_len` | Read length in bp |
| `status` | QC call (see status codes below) |
| `assigned_sh` | Matched shRNA name from the design |
| `assigned_oligo_well` | Corresponding well in the design plate |
| `bc_found` | Barcode extracted from the read |
| `exp_bc_in_read` | Expected barcode in read orientation (RC of design BC) |
| `bc_mm` | Mismatches between found and expected BC |
| `bc_mutations` | Position and base change of each BC mismatch |
| `sense_found` | Sense sequence extracted from the read |
| `sense_mm` | Mismatches in sense region |
| `sense_mutations` | Position and base change of each sense mismatch |
| `anti_found` | Antisense sequence extracted from the read |
| `anti_mm` | Mismatches in antisense region |
| `anti_mutations` | Position and base change of each antisense mismatch |
| `uci` | UCI sequence extracted from the read |
| `hairpin_ok` | TRUE if sense == RC(antisense) |
| `min_q_bc` | Minimum Phred quality in BC region (requires ABI ZIP) |
| `mean_q_bc` | Mean Phred quality in BC region |
| `min_q_sense` | Minimum Phred quality in sense region |
| `mean_q_sense` | Mean Phred quality in sense region |
| `anchor_mm` | Mismatches in anchor sequence |
| `notes` | Additional flags (MluI mutation, BC/shRNA discordance) |

---

## Status codes

| Status | Meaning | Action |
|--------|---------|--------|
| `PASS` | BC and shRNA both match perfectly (0 mm each) | Use this colony |
| `PASS (1bp BC mut)` | shRNA perfect, 1 mismatch in BC | Likely OK — check `bc_mutations` column; if not in a critical position, use the colony |
| `PASS (2bp BC mut)` | shRNA perfect, 2 mismatches in BC | Caution — verify that the BC still uniquely identifies this shRNA before using |
| `PASS (1bp shRNA mut)` | BC perfect, 1 total shRNA mismatch | Likely OK — check `sense_mutations` / `anti_mutations`; if not in the seed region, may still be functional |
| `PASS (2bp shRNA mut)` | BC perfect, 2 total shRNA mismatches | Caution — two mutations in the shRNA may affect silencing efficiency; sequence-verify before using |
| `CHECK` | BC matches (0 mm) but shRNA has more mismatches than threshold | Inspect the actual sequence manually — could be a sequencing artefact in the hairpin region |
| `INVALID HAIRPIN` | Sense ≠ RC(antisense) | The hairpin structure is broken; do not use this colony. Check sequencing quality in the shRNA region (`min_q_sense`) |
| `shRNA OK / BC WRONG` | shRNA matches perfectly but BC has > 2 mm | Possible BC synthesis error or barcode swap; do not use without further investigation |
| `BC/shRNA discordant` | BC and shRNA each match, but to different library members | Possible chimeric colony or mixed ligation; do not use |
| `SEQ TRUNCATED` | Anchor sequence not found in the read | The read may be too short to reach the insert, or the insert was not cloned; pick another colony |
| `SEQ FAILED` | Read shorter than 150 bp | Sequencing failed; re-sequence or pick another colony |

---

## Adjusting matching thresholds

By default, a well is called `PASS` with up to 2 mismatches in the BC and
up to 4 total mismatches in the shRNA (sense + antisense combined). You can
tighten or relax these:

```{r thresholds}
# Strict: only perfect matches
result_strict <- catcheR_sangerQC(
  fasta         = "11109801799-1.fasta",
  design_xlsx   = "EB003C_DES01_shRNA_tdark_Ligation.xlsx",
  output_dir    = "Output/",
  sample_name   = "plate1_strict",
  bc_mm_pass    = 0L,    # no BC mismatches allowed for PASS
  shrna_mm_pass = 0L     # no shRNA mismatches allowed for PASS
)

# Lenient: allow more mismatches (useful for very short reads)
result_lenient <- catcheR_sangerQC(
  fasta         = "11109801799-1.fasta",
  design_xlsx   = "EB003C_DES01_shRNA_tdark_Ligation.xlsx",
  output_dir    = "Output/",
  sample_name   = "plate1_lenient",
  bc_mm_pass    = 2L,
  shrna_mm_pass = 6L
)
```

---

## Using quality scores to interpret borderline results

When you provide the ABI ZIP file, the columns `min_q_bc`, `mean_q_bc`,
`min_q_sense`, and `mean_q_sense` give you the Phred quality scores in the
extracted regions. A Phred score of 30 or above (error rate < 0.1%) is
generally considered reliable. A score below 20 (error rate > 1%) in the
BC region should make you sceptical of mismatches reported there.

```{r quality}
# Wells where BC mismatches may be sequencing errors (low quality)
result$results[
  result$results$bc_mm > 0 &
  !is.na(result$results$min_q_bc) &
  result$results$min_q_bc < 20,
  c("sanger_well","assigned_sh","bc_mm","min_q_bc","bc_mutations","status")
]
```

---

## Next steps

After confirming correct colonies by Sanger QC:

1. Pick and expand the colonies that passed.
2. Complete the final cloning step (transfer insert into final backbone).
3. Run `catcheR_step1QC()` on the intermediate pool FASTQ to detect any
   barcode–shRNA swaps at scale across the full library.
4. Proceed to `catcheR_step2QC_extraction()` after the final plasmid pool
   is sequenced.

---

## Full parameter reference

```{r params}
catcheR_sangerQC(
  fasta           = "sequences.fasta",   # FASTA with Sanger reads
  design_xlsx     = "design.xlsx",       # Design Excel with TOP OLIGO column
  output_dir      = "Output/",           # Root output directory
  sample_name     = NULL,                # Optional label added to run folder
  plate_sheet     = "Plate 01",          # Excel sheet name
  oligo_col       = 2L,                  # 0-based column index of oligo name
  sequence_col    = 4L,                  # 0-based column index of TOP OLIGO
  first_row       = 8L,                  # 0-based row of first oligo
  last_row        = 67L,                 # 0-based row of last oligo
  ligation_split  = 6L,                  # Column split between ligation conditions
  ligation_labels = c("1:12", "1:20"),   # Names of the two ligation conditions
  anchor          = "GGCGCGCCATTTAAATGTCGAC",  # MCS anchor in the read
  anchor_mm       = 2L,                  # Max mismatches in anchor search
  bc_mm_pass      = 2L,                  # Max BC mm for PASS call
  shrna_mm_pass   = 4L,                  # Max shRNA total mm for PASS call
  min_read_len    = 150L,                # Minimum read length to attempt extraction
  abi_zip         = NULL,                # Optional ZIP with PHD quality files
  phd_prefix      = NULL,                # Auto-detected from ZIP contents
  sh_id_prefix    = "siKDBC_"            # Prefix stripped from oligo names
)
```
