---
title: "catcheR 10X QC, filtercatch, and nocatch workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{catcheR 10X QC, filtercatch, and nocatch workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Purpose

This vignette describes the modular post-catching functions:

- `catcheR_10XcatchQC()`: explores barcode/UMI distributions and writes a UMI threshold.
- `catcheR_filtercatch()`: filters caught cells using the UMI threshold and expression matrix.
- `catcheR_nocatch()`: processes no-catch/empty-cell information and optionally merges samples.

Use these functions when you want more control than the all-in-one `catcheR_10Xcatch()` wrapper.

# Input folder

After the catching/slicing step, the working folder should contain the intermediate files expected by the original scripts, for example:

```text
scratch/10X/
  Results_1/
  rc_barcodes_genes.csv
  filtered_feature_bc_matrix.h5
```

# 1. 10X catching QC

```{r run-10xcatchqc}
library(catcheR)

catch_qc <- catcheR_10XcatchQC(
  folder      = "scratch/10X/",
  output_dir  = "Output/",
  sample_name = "AB017",
  reference   = "GGCGCGTTCATCTGGGGGAGCCG",
  mode        = "bimodal",
  samples     = 1,
  x           = 100,
  y           = 400
)
```

The key checkpoint file is:

```text
UMI_threshold.txt
```

It is copied into the run's `to_scratch/` folder so that the filtering step can be resumed later.

# 2. Filter caught cells

```{r run-filtercatch}
filtered <- catcheR_filtercatch(
  folder            = "scratch/10X/",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  output_dir        = "Output/",
  sample_name       = "AB017",
  UMI.count         = NULL,
  percentage        = 15,
  ratio             = 5,
  samples           = 1
)
```

If `UMI.count = NULL`, the function tries to read the threshold from `UMI_threshold.txt` generated by `catcheR_10XcatchQC()`.

Important output files include silencing matrices copied to:

```text
Output/filtercatch/YYMMDD_filtercatch_AB017/to_scratch/
```

# 3. Process no-catch / empty selection

```{r run-nocatch}
nocatch <- catcheR_nocatch(
  folder            = "scratch/10X/",
  expression.matrix = "filtered_feature_bc_matrix.h5",
  output_dir        = "Output/",
  sample_name       = "AB017",
  threshold         = NULL,
  samples           = 1,
  ref               = "TACGCGTTCATCTGGGGGAGCCG",
  merge.samples     = TRUE
)
```

If `merge.samples = TRUE`, the function also runs the all-sample merging script.

# Output folders

Each modular function writes a dated run folder:

```text
Output/10XcatchQC/YYMMDD_10XcatchQC_AB017/
Output/filtercatch/YYMMDD_filtercatch_AB017/
Output/nocatch/YYMMDD_nocatch_AB017/
```

Each folder contains:

```text
csv/
plots/
stats/
R_objects/
to_scratch/
```

# Returned R objects

```{r inspect-filter-nocatch}
names(catch_qc)
catch_qc$paths

names(filtered)
filtered$paths

names(nocatch)
nocatch$paths
```

# Restart from checkpoint

```bash
# Restart filtering from 10XcatchQC output
cp Output/10XcatchQC/YYMMDD_10XcatchQC_AB017/to_scratch/* scratch/10X/

# Restart nocatch from filtercatch output
cp Output/filtercatch/YYMMDD_filtercatch_AB017/to_scratch/* scratch/10X/
```
